The Proofs

### 2023-11-13 proofs/tex/decl-typing-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Restructure proofs

\begin{lemma} \label{lemma:app-inf-equ-stable}
    If $[[Γ; Φ ⊢ iN1 ● args ⇒> iM]]$ and $[[Γ ⊢ iN1 ≈ iN2]]$ 
    then $[[Γ; Φ ⊢ iN2 ● args ⇒> iM]]$.
\end{lemma}
\begin{proof}
    By \cref{lemma:equiv-completeness}, 
    $[[Γ ⊢ iN1 ≈ iN2]]$ implies $[[iN1 ≈ iN2]]$.
    Let us prove the required judgement by induction on $[[iN1 ≈ iN2]]$.
    Let us consider the last rule used in the derivation.
    \begin{caseof}
        \item \ruleref{\ottdruleEOneNVarLabel}.
            It means that $[[iN1]]$ is $[[α⁻]]$ and $[[iN2]]$ is $[[α⁻]]$.
            Then the required property coincides with the assumption. 
        \item \ruleref{\ottdruleEOneShiftULabel}. 
            It means that $[[iN1]]$ is $[[↑iP1]]$ and $[[iN2]]$ is $[[↑iP2]]$.
            where $[[iP1 ≈ iP2]]$.

@@

            By inversion of $[[Γ; Φ ⊢ (iQ1 → iN1) ● v, args ⇒> iM]]$:
            \begin{enumerate}
                \item $[[Γ ; Φ ⊢ v : iP]]$
                \item $[[Γ ⊢ iQ1 ≥ iP]]$,
                    and then by transitivity \cref{corollary:subtyping-transitivity},
                    $[[Γ ⊢ iQ2 ≥ iP]]$;
                \item $[[Γ ; Φ ⊢ iN1 ● args ⇒> iM]]$, 
                    and then by induction hypothesis, $[[Γ ; Φ ⊢ iN2 ● args ⇒> iM]]$.
            \end{enumerate}

@@

\newcommand{\pureSize}[1]{\ensuremath{\mathsf{pure\_size}(#1)}}
\newcommand{\metric}[1]{\ensuremath{\mathsf{metric}(#1)}}
\newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
\newcommand{\npq}[1]{\ensuremath{\mathsf{npq}(#1)}}

\begin{definition}[Number of prenex quantifiers]
    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
    \begin{itemize}
        \item [$+$] $\npq{[[∃nas.iP]]} = |nas|$, if $[[iP ≠ ∃nbs.iP']]$,
        \item [$-$] $\npq{[[∀pas.iN]]} = |pas|$, if $[[iN ≠ ∀pbs.iN']]$.
    \end{itemize}
\end{definition}

\begin{definition}[Size of a Declarative Judgement]
    \label{def:decl-typing-size}
    For a declarative typing judgement $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
            (\size{[[args]]}, \npq{[[iN]]})$)
    \end{itemize}
    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
    syntax tree of the term $[[v]]$ or $[[c]]$
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
\end{definition}

\begin{definition}[Number of Equivalence Nodes]
    For a tree $T$ inferring
    a declarative typing judgement,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

\begin{definition}[Metric]
    \label{def:decl-typing-metric}
    For a tree $T$ inferring
    a declarative typing judgement $J$,
    let us define a metric $\metric{T}$
    as a pair $(\size{J}, \eqNodes{T})$.
\end{definition}

\begin{lemma}[Declarative typing is preserved under context equivalence]
    \label{lemma:decl-typing-context-equiv}
    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
    \begin{itemize}
        \item [$+$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$.
        \item [$-$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$.
        \item [$\bullet$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us prove it by induction on the $\metric{T_1}$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

>>>>

\begin{definition}[Number of prenex quantifiers]
    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
    \begin{itemize}
        \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
        \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
    \end{itemize}
\end{definition}

\begin{definition}[Size of a Declarative Judgement]
    \label{def:decl-typing-size}
    For a declarative typing judgment $J$
    let us define a metrics $\size{J}$ as a pair of numbers 
    in the following way:
    \begin{itemize}
        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
            (\size{[[args]]}, \npq{[[iN]]})$)
    \end{itemize}
    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
    syntax tree of the term $[[v]]$ or $[[c]]$
    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
\end{definition}

\begin{definition}[Number of Equivalence Nodes]
    For a tree $T$ inferring
    a declarative typing judgment,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

\begin{definition}[Metric]
    \label{def:decl-typing-metric}
    For a tree $T$ inferring
    a declarative typing judgment $J$,
    let us define a metric $\metric{T}$
    as a pair $(\size{J}, \eqNodes{T})$.
\end{definition}

\lemmaAppInfEquStable*
\begin{proof}
    By \cref{lemma:equiv-completeness}, 
    $[[Γ ⊢ iN1 ≈ iN2]]$ implies $[[iN1 ≈ iN2]]$.
    Let us prove the required judgement by induction on $[[iN1 ≈ iN2]]$.
    Let us consider the last rule used in the derivation.
    \begin{caseof}
        \item \ruleref{\ottdruleEOneNVarLabel}.
            It means that $[[iN1]]$ is $[[α⁻]]$ and $[[iN2]]$ is $[[α⁻]]$.
            Then the required property coincides with the assumption. 
        \item \ruleref{\ottdruleEOneShiftULabel}. 
            It means that $[[iN1]]$ is $[[↑iP1]]$ and $[[iN2]]$ is $[[↑iP2]]$.
            where $[[iP1 ≈ iP2]]$.

@@

            By inversion of $[[Γ; Φ ⊢ (iQ1 → iN1) ● v, args ⇒> iM]]$:
            \begin{enumerate}
                \item $[[Γ ; Φ ⊢ v : iP]]$
                \item $[[Γ ⊢ iQ1 ≥ iP]]$,
                    and then by transitivity \cref{lemma:subtyping-transitivity},
                    $[[Γ ⊢ iQ2 ≥ iP]]$;
                \item $[[Γ ; Φ ⊢ iN1 ● args ⇒> iM]]$, 
                    and then by induction hypothesis, $[[Γ ; Φ ⊢ iN2 ● args ⇒> iM]]$.
            \end{enumerate}

@@

\lemmaDeclTypingContextEquiv*
\begin{proof}
    Let us prove it by induction on the $\metric{T_1}$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

### 2023-11-13 proofs/tex/equiv-alg-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Restructure proofs

\begin{lemma}[Algorithmization of declarative equivalence]
  \label{lemma:decl-equiv-algorithmization}
  Declarative equivalence is the equality of normal forms. 
  \begin{itemize}
    \item[$+$] $[[iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
    \item[$-$] $[[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
  \end{itemize}
\end{lemma}
\begin{proof} \hfill
  \begin{itemize}
    \item[$+$] Let us prove both directions separately.
    \begin{itemize}
      \item[$\Rightarrow$] 
        exactly by \cref{lemma:normalization-completeness},
      \item[$\Leftarrow$] 
        from \cref{lemma:normalization-soundness}, we know
        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
        $[[iP ≈ iQ]]$.
    \end{itemize}
    \item[$-$] The proof is exactly the same.
  \end{itemize}
\end{proof}

\begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
  \label{lemma:wf-equiv-algo}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof coincides with the proof of \cref{lemma:wf-equiv},
  and adds two cases for equating two positive or two negative algorithmic variables,
  which must be equal by inversion, and thus, 
  $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
\end{proof}

\begin{corollary}[Normalization preserves well-formedness of algorithmic types]
  \label{corollary:wf-nf-algo}
  \begin{itemize}
    \item[$+$] $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ nf(uP)]]$,
    \item[$-$] $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ nf(uN)]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:wf-equiv-algo,lemma:normalization-soundness-alg}.
\end{proof}

\begin{corollary}[Normalization preserves the signature of the algorithmic substitution]
  \label{corollary:norm-subst-sig-algo}
  \hfill \\
  $[[Θ ⊢ uσ : Ξ]] \iff [[Θ ⊢ nf(uσ) : Ξ]]$,
  $[[Γ ⊢ uσ : Ξ]] \iff [[Γ ⊢ nf(uσ) : Ξ]]$.
\end{corollary}
\begin{proof}
  The proof is analogous to \cref{corollary:wf-s-nf}.
\end{proof}

\begin{lemma}[Algorithmic substitution equivalence becomes equality after normalization]
  \label{corollary:subst-subt-equiv-algorithmization}
  Suppose that $[[Θ ⊢ uσ1 : Ξ']]$ and $[[Θ ⊢ uσ2 : Ξ']]$ are algorithmic substitutions
  and $[[Ξ ⊆ Ξ']]$. 
  Then $[[Θ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
\end{lemma}
\begin{proof}
  Follows immediately from \cref{lemma:subt-equiv-algorithmization}:
  \begin{itemize}
    \item [$\Rightarrow$]
      If $[[α̂± ∉ Ξ]]$, then $[[ [nf(uσ1)|Ξ]α̂± = [nf(uσ2)|Ξ]α̂± ]] = [[α̂±]]$ by 
      definition. 
      For any $[[α̂± ∊ Ξ]]$, 
      $[[ [nf(uσ1)|Ξ]α̂± = nf([uσ1]α̂±)]]$ and 
      $[[ [nf(uσ2)|Ξ]α̂± = nf([uσ2]α̂±) ]]$; 
      $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ implies
      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by \cref{lemma:decl-equiv-algorithmization}.
    \item [$\Leftarrow$]
      If $[[α̂± ∊ Ξ]]$, then 
      $[[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$ implies
      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by definition 
      of substitution restriction and normalization.
      In turn, $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ means
       $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ by \cref{lemma:decl-equiv-algorithmization}.
  \end{itemize}
\end{proof}

>>>>

\lemmaWfEquivAlgo*
\begin{proof}
  The proof coincides with the proof of \cref{lemma:wf-equiv},
  and adds two cases for equating two positive or two negative algorithmic variables,
  which must be equal by inversion, and thus, 
  $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
\end{proof}

\corollaryWfNfAlgo*
\begin{proof}
  Immediately from \cref{lemma:wf-equiv-algo,lemma:normalization-soundness-alg}.
\end{proof}

\corollaryNormSubstSigAlgo*
\begin{proof}
  The proof is analogous to \cref{corollary:wf-s-nf}.
\end{proof}

\corollarySubstSubtEquivAlg*
\begin{proof}
  Follows immediately from \cref{lemma:subt-equiv-algorithmization}:
  \begin{itemize}
    \item [$\Rightarrow$]
      If $[[α̂± ∉ Ξ]]$, then $[[ [nf(uσ1)|Ξ]α̂± = [nf(uσ2)|Ξ]α̂± ]] = [[α̂±]]$ by 
      definition. 
      For any $[[α̂± ∊ Ξ]]$, 
      $[[ [nf(uσ1)|Ξ]α̂± = nf([uσ1]α̂±)]]$ and 
      $[[ [nf(uσ2)|Ξ]α̂± = nf([uσ2]α̂±) ]]$; 
      $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ implies
      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by \cref{lemma:decl-equiv-algorithmization}.
    \item [$\Leftarrow$]
      If $[[α̂± ∊ Ξ]]$, then 
      $[[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$ implies
      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by definition 
      of substitution restriction and normalization.
      In turn, $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ means
       $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ by \cref{lemma:decl-equiv-algorithmization}.
  \end{itemize}
\end{proof}
